Numerical and asymptotic solution of nonlinear electrokinetic phenomena

### 2012-04-02 notes/article/main.tex
Work on article

\usepackage{setspace}\doublespacing

@@

\begin{abstract}
\end{abstract}

\begin{keyword}
numerical, solver, electrokinetics, nonlinear
\end{keyword}

@@

Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge, thus
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle's
size, an analytical asymptotic solution for the Debye layer
dynamics can be found \cite{yariv2010asymptotic}.

@@

A closed-form linear asymptotic solution has been developed for
spherical ion-exchanging particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena are expected.
This interesting regime has not yet been explored.
Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects in many nanotechnology applications,
such as the self-assembly of superstructures, roving sensors, 
drug-delivery systems and useful nanomachinery 
\cite{howse2007self,paxton2004catalytic,pumera2010electrochemically}.
However, since the model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver for this system is of importance.

The goal of this work is to develop and implement an iterative numerical
solver for nonlinear electrokinetic problems, and apply it to the study
of systems that have no closed-form solutions.
This solver can be used to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also to possibly lead to further analytical developments, based on new scaling behavior
discovered numerically. We may also target interesting physical
phenomena that have been observed experimentally.

@@

The cubic terms $\bx_3$ are computed by solving a linear PDE system 
with a right-hand side determined by the linear terms $\bx_1$ and the quadratic terms $\bx_2$:
\begin{eqnarray*}
\varPhi_3 &=& \cos\theta \left(\frac{15 \cU_1\alpha}{64} + \frac{3}{32}\right) - \frac{3 \cU_1\alpha {\cos}^3\theta}{32} - \frac{\cos\theta \left( - 183 {\cU_1}^2\alpha^2 + 839 \cU_1\alpha + 470\right)}{2560 r^2} \\ 
&&+ \frac{3 \cos\theta \left(5 \cU_1\alpha - 6 {\cos^2\theta} - 4 \cU_1\alpha {\cos^2\theta} + 4\right)}{64 r^3} + \frac{\left(2 \cU_1\alpha - 1\right) \left(\cos\theta - 3 {\cos}^3\theta\right)}{64 r^5} 
\\ 
&&- \frac{\left(3 \cos\theta - 5 {\cos}^3\theta\right) \left( - \frac{19 {\cU_1}^2\alpha^2}{5120} + \frac{97 \cU_1\alpha}{5120} + \frac{3 \cU_2\alpha}{320} + \frac{21}{1280}\right)}{r^4} + \frac{{\cos}^3\theta \left(\frac{15 \cU_1\alpha}{64} + \frac{3}{32}\right)}{r^2} \\ 
&&+ \frac{{\cos}^3\theta \left(\frac{\cU_1\alpha}{64} + \frac{3}{64}\right)}{r^6},
\\
C_3 &=&
\frac{3 {{\cU_1}}^2\alpha^2 {\cos^3\theta}}{32} + \frac{\cos\theta \left(797 {{\cU_1}}^2\alpha^2 + 419 {\cU_1}\alpha - 512 {\cU_2}\alpha + 174\right)}{2560 r^2} \\
&&- \frac{\left(3 \cos\theta - 5 {\cos^3\theta}\right) \left(\frac{59 {{\cU_1}}^2\alpha^2}{5120} + \frac{103 {\cU_1}\alpha}{5120} + \frac{69 {\cU_2}\alpha}{320} + \frac{3}{1280}\right)}{r^4} \\
&&- \frac{3\alpha \cos\theta \left( - 8\alpha {{\cU_1}}^2 {\cos^2\theta} + 7\alpha {{\cU_1}}^2 + 2 {\cU_1} + 32 {\cU_2} {\cos^2\theta} - 32 {\cU_2}\right)}{128 r^3} - \frac{3 {\cU_1}\alpha \cos\theta \left(5 {\cU_1}\alpha + 2\right)}{64} \\
&&+ \frac{\alpha \left(\cos\theta - 3 {\cos^3\theta}\right) \left(5\alpha {{\cU_1}}^2 + 2 {\cU_1} + 16 {\cU_2}\right)}{128 r^5} + \frac{{{\cU_1}}^2\alpha^2 {\cos^3\theta}}{32 r^6} - \frac{3 {\cU_1}\alpha {\cos^3\theta} \left(5 {\cU_1}\alpha + 2\right)}{64 r^2},
\\
\Psi_3 &=&
r^2 {\sin^2\theta} \left(\frac{{\cW_2}}{15} - \frac{{\cW_1}}{3} + \frac{209}{3360}\right) 
 - \frac{{\sin^2\theta} \left(\frac{5 {\cW_2}}{3} - \frac{{\cW_1}}{3} + \frac{35}{264}\right)}{r}
\\ && + \frac{{\sin^4\theta} \left(2 {\cW_2} + \frac{761}{5632}\right)}{r} - \frac{{\sin^2\theta} \left(1848 r^6 {\sin^2\theta} - 2079 r^3 {\sin^2\theta} + 924 r^3 - 7 {\sin^2\theta} + 10\right)}{19712 r^5} 
\\ && + \frac{\left(\frac{2 {\cW_2}}{5} + \frac{829}{28160}\right) \left(4 {\sin^2\theta} - 5 {\sin\theta}^4\right)}{r^3} + \frac{{\sin^2\theta} \left(\frac{{\cU_1}\alpha}{16} - \frac{1}{8}\right) \left(2 r^3 - 3 r^2 + 1\right)}{4 r},
\\
\cW_1 &=& \frac{69}{512 \left(\sqrt{\gamma} + 1\right)} - \frac{1407\alpha^2 {\log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}^3}{1280} - \frac{123 \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{1280} - \frac{27}{512 {\left(\sqrt{\gamma} + 1\right)}^2} \\ && -
\frac{1899\alpha {\log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}^2}{2560} - \frac{\alpha \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right) \left(\frac{81}{16 \left(\sqrt{\gamma} + 1\right)} - \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right) \left(\frac{27\alpha \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{4} + \frac{27}{8}\right)\right)}{320} \\ && + 
\frac{21\alpha \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{128 \left(\sqrt{\gamma} + 1\right)},
\\
\cW_2 &=& \frac{9 \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{256} + \frac{21\alpha^2 {\log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}^3}{64} - \frac{27}{512 \left(\sqrt{\gamma} + 1\right)} - \frac{27}{512 {\left(\sqrt{\gamma} + 1\right)}^2}
 + \frac{15\alpha {\log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}^2}{64} 
\\ && 
- \frac{297\alpha \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{1024 \left(\sqrt{\gamma} + 1\right)}.
\end{eqnarray*}

@@

Moreover, two interesting phenomena (cubic regime and downstream vortex) were predicted
using the numerical results, and later verified using asymptotic expansion of the
nonlinear system.

>>>>

%\usepackage{setspace}\doublespacing

@@

\begin{abstract}
A numerical solver for the simulation of ion-exchanging particles migration in an
electrolyte solution is presented. The electrokinetic transport process is
described by a nonlinear partial differential equation system, 
as well as nonlinear boundary conditions.
The nonlinear system is discretized and solved iteratively using Newton's method.
An asymptotic analytical solution is derived for the validation of the solver. 
The numerical results are compared to the asymptotic solutions, and good correspondence is achieved.
\end{abstract}

\begin{keyword}
CFD \sep electrokinetics \sep asymptotics
\end{keyword}

@@

Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge, thus
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle's
size, an analytical asymptotic solution for the Debye layer
dynamics can be found \cite{yariv2010asymptotic}.

@@

A closed-form linear asymptotic solution has been developed for
spherical ion-exchanging particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena are expected.
This interesting regime has not yet been explored.

Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects in many nanotechnology applications,
such as the self-assembly of superstructures, roving sensors, 
drug-delivery systems and useful nanomachinery 
\cite{howse2007self,paxton2004catalytic,pumera2010electrochemically}.

However, since the model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver for this system is of importance.

The goal of this work is to develop and implement an iterative numerical
solver for nonlinear electrokinetic problems, and apply it to the study
of systems that have no closed-form solutions.
This solver can be used to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also to possibly lead to further analytical developments, based on new scaling behavior
discovered numerically. We may also target interesting physical
phenomena that have been observed experimentally.
One such phenomenon is nonlinear electrokinetic vortex flow (as shown in Figure \ref{fig:Granule}), 
which can useful for microfluidic mixing applications \cite{wang2004mix, ben2002vortex},

\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=0.4\textwidth]
            {figs/vortex.eps}
        \caption{A micro-vortex electrokinetic flow around a 1 mm spherical ion-exchanger granule. }
        \label{fig:Granule}
    \end{center}
\end{figure}

@@

The cubic terms $\bx_3$ are computed by solving a linear PDE system 
with a right-hand side determined by the linear terms $\bx_1$ and the quadratic terms $\bx_2$:
\begin{eqnarray*}
\varPhi_3 &=& \cos\theta \left(\frac{15 \cU_1\alpha}{64} + \frac{3}{32}\right) - \frac{3 \cU_1\alpha {\cos}^3\theta}{32} - \frac{\cos\theta \left( - 183 {\cU_1}^2\alpha^2 + 839 \cU_1\alpha + 470\right)}{2560 r^2} \\ 
&&+ \frac{3 \cos\theta \left(5 \cU_1\alpha - 6 {\cos^2\theta} - 4 \cU_1\alpha {\cos^2\theta} + 4\right)}{64 r^3} + \frac{\left(2 \cU_1\alpha - 1\right) \left(\cos\theta - 3 {\cos}^3\theta\right)}{64 r^5} 
\\ 
&&- \frac{\left(3 \cos\theta - 5 {\cos}^3\theta\right)}{r^4} 
\left( - \frac{19 {\cU_1}^2\alpha^2}{5120} + \frac{97 \cU_1\alpha}{5120} + 
\frac{3 \cU_2\alpha}{320} + \frac{21}{1280}\right) \\
&& + \frac{{\cos}^3\theta}{r^2} \left(\frac{15 \cU_1\alpha}{64} + \frac{3}{32}\right)  
+ \frac{{\cos}^3\theta}{r^6} \left(\frac{\cU_1\alpha}{64} + \frac{3}{64}\right),
\\
C_3 &=&
\frac{3 {{\cU_1}}^2\alpha^2 {\cos^3\theta}}{32} + \frac{\cos\theta \left(797 {{\cU_1}}^2\alpha^2 + 419 {\cU_1}\alpha - 512 {\cU_2}\alpha + 174\right)}{2560 r^2} \\
&&- \left(\frac{59 {{\cU_1}}^2\alpha^2}{5120} + \frac{103 {\cU_1}\alpha}{5120} + \frac{69 {\cU_2}\alpha}{320} + \frac{3}{1280}\right)\frac{\left(3 \cos\theta - 5 {\cos^3\theta}\right)}{r^4} \\
&&- \frac{3\alpha \cos\theta \left( - 8\alpha {{\cU_1}}^2 {\cos^2\theta} + 7\alpha {{\cU_1}}^2 + 2 {\cU_1} + 32 {\cU_2} {\cos^2\theta} - 32 {\cU_2}\right)}{128 r^3} - \frac{3 {\cU_1}\alpha \cos\theta \left(5 {\cU_1}\alpha + 2\right)}{64} \\
&&+ \frac{\alpha \left(\cos\theta - 3 {\cos^3\theta}\right) \left(5\alpha {{\cU_1}}^2 + 2 {\cU_1} + 16 {\cU_2}\right)}{128 r^5} + \frac{{{\cU_1}}^2\alpha^2 {\cos^3\theta}}{32 r^6} - \frac{3 {\cU_1}\alpha {\cos^3\theta} \left(5 {\cU_1}\alpha + 2\right)}{64 r^2},
\\
\Psi_3 &=&
r^2 {\sin^2\theta} \left(\frac{{\cW_2}}{15} - \frac{{\cW_1}}{3} + \frac{209}{3360}\right) 
 - \frac{{\sin^2\theta}}{r} \left(\frac{5 {\cW_2}}{3} - \frac{{\cW_1}}{3} + \frac{35}{264}\right)
\\ && + \frac{{\sin^4\theta} \left(2 {\cW_2} + \frac{761}{5632}\right)}{r} - \frac{{\sin^2\theta} \left(1848 r^6 {\sin^2\theta} - 2079 r^3 {\sin^2\theta} + 924 r^3 - 7 {\sin^2\theta} + 10\right)}{19712 r^5} 
\\ && + \frac{\left(\frac{2 {\cW_2}}{5} + \frac{829}{28160}\right) \left(4 {\sin^2\theta} - 5 {\sin^4\theta}\right)}{r^3} + \frac{{\sin^2\theta} \left(\frac{{\cU_1}\alpha}{16} - \frac{1}{8}\right) \left(2 r^3 - 3 r^2 + 1\right)}{4 r},
\\
\cW_1 &=& \frac{69}{512 \left(\sqrt{\gamma} + 1\right)} - \frac{1407\alpha^2 {\log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}^3}{1280} - \frac{123 \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{1280} - \frac{27}{512 {\left(\sqrt{\gamma} + 1\right)}^2} \\ && -
\frac{1899\alpha {\log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}^2}{2560} - \frac{\alpha \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right) \left(\frac{81}{16 \left(\sqrt{\gamma} + 1\right)} - \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right) \left(\frac{27\alpha \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{4} + \frac{27}{8}\right)\right)}{320} \\ && + 
\frac{21\alpha \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{128 \left(\sqrt{\gamma} + 1\right)},
\\
\cW_2 &=& \frac{9 \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{256} + \frac{21\alpha^2 {\log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}^3}{64} - \frac{27}{512 \left(\sqrt{\gamma} + 1\right)} - \frac{27}{512 {\left(\sqrt{\gamma} + 1\right)}^2}
 + \frac{15\alpha {\log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}^2}{64} 
\\ && 
- \frac{297\alpha \log\left(\frac{\sqrt{\gamma} + 1}{2 \sqrt{\gamma}}\right)}{1024 \left(\sqrt{\gamma} + 1\right)}.
\end{eqnarray*}

@@

Moreover, two interesting phenomena (cubic regime and downstream vortex) were predicted
using the numerical results, and were verified using asymptotic expansion of the
nonlinear system.
